Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-01-04 paper/paper.tex
fix typos

\begin{abstract}
We introduce a method of using generic programming tensor network to compute various properties of independent sets,
which includes: maximum size of such sets, the number of sets of a given size, enumeration of sets of a given size.
%Although the computational complexity inevitably scales exponentially with treewidth of the graph,
%our method using tensor network contraction with generic programming is highly versatile and is able to compute properties that are otherwise not known to be possible for up to several hundred vertices.
At the same time, our algorithms are very simple to implement and one can directly utilize recent advances in tensor network contraction techniques like near optimal contraction order finding and slicing.
%One can easily find a good contraction order for the tensor network representation of a graph $G$ with thousands of vertices. 
The algorithmic complexity of this approach is $2^{{\rm tw}(G)}$, where ${\rm tw} (G)$ the tree width of the problem graph.
Our generic programming tensor network method can be used to compute properties of other problems like cut size, coloring, maximal cliques et al.
To demonstrate the versatility of this tool, we apply it to a few examples including the calculations of the hard-square entropy constant and overlap gap property of three regular graphs.
\end{abstract}

@@

\section{Introduction}
In graph theory and combinatorial optimization, there are many interesting hard computational problems concerning properties of independent sets.
%An independent set of a graph is a subset of its vertices that no two vertices in this set are adjacent to each other.
For an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, $(u, v) \not\in E$. %there is no edge connecting $u$ and $v$ in $G$.
The primary insterest of many computational complexity theorists is finding the maximum size of such vertex sets $\alpha(G) \equiv \max_{I}|I|$ and one of the sets of this size.
Finding $\alpha(G)$ is a hard computational problem; moreover, it is NP-hard to even approximate $\alpha(G)$ within $|V|^{1-\epsilon}$ ~\cite{Hastad1996},
meaning there is unlikely a polynomial time algorithm to find an independent set with size $\alpha(G)/|V|^{1-\epsilon}$ for an arbitrarily small positive $\epsilon$.
Naive exhaustive search for an MIS requires computing time $O\left(2^{|V|} \right)$.
People developed more efficient exact algorithms for finding MISs, including the branching algorithms~\cite{Tarjan1977,Robson1986} and dynamic programming.
The branching algorithms can reduce the base of the exponential time scaling to, e.g., $1.1996^{|V|} {|V|}^{O(1)}$~\cite{Xiao2017}, 
while dynamic programming approaches~\cite{Courcelle1990, Fomin2013} works better for graphs with a small treewidth $\text{tw}(G)$, and
these methods can produce algorithms of complexity $O(2^{\text{tw}(G)} \text{tw}(G) |V|)$.
People are eager to find better algorithms for solving the MIS problem;
not only for it has a wide range of applications in scheduling, logistics, social network analysis, bioinformatics, wireless networks and telecommunication,
map labelling, computer vision, etc.~\cite{Butenko2003, Wu2015}, but also for it is a well know NP-complete problem that can be mapped from
 many other important combinatorial optimization problems in polynomial time such as 3-satisfiability problem, the maximum clique problem and the maximum vertex cover problem~\cite{Moore2011}.

%more concretely, an maximal independent set of a graph $G$ corresponds to a maximum clique in the complement graph of $G$,
%and the complement vertex set of an MIS corresponds to a minimum vertex cover of the graph $G$.
%In this paper, the term ``maximal'' has a different meaning from ``maximum''  which we discussed above; a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.
In this paper, we do not limit our discussion to finding one of the maximum indepedent sets and its size.
There are many other interesting problems pertaining to independent sets, such as the number of independent sets of a given size
and enumerating independent sets at a give size or what we call computing \textit{properties}.
Some of these problems are of great interest in physics applications such as the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018} \red{cite experiment when it's ready};
they can, for example, be used to understand phase transitions, to analyse the overlap gap property~\cite{Gamarnik2013, Gamarnik2019} and to identify harder graphs in an ensemble of graphs \red{cite experiment}.
Finding the number of independent sets of a given size is equivalent to computing the coefficients of independence polynomial,
which is a useful graph characteristics that can related to Euler characteristics of the independence complex~\cite{Bousquet2008,Levit2009}, partition functions~\cite{Lee1952,Yang1952} et al.
Computing independence polynomials belongs to the complexity class \#P-hard 
and only exponential algorithms~\cite{Ferrin2014} are known to compute it exactly.
There are also some interesting works on efficiently approximating the independence polynomial~\cite{Harvey2018}, but in this work, we mainly focus on computing the independence polynomial exactly.
We introduce a tensor-network based framework to compute the various properties pertaining to independent sets.
We map the independent set problem into generic tensor network contraction with specially designed tensor element algebra.
Its algorithmic complexity is similar to that of dynamic programming, i.e. scales exponentially as ${\rm tw}(G)$.
%But the tensor network approach is highly versatile so different problems can be generically solved with minimal implementation efforts without loss of efficiency.
We benchmark our algorithms by computing various independent set properties on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance on sparse graphs.
Its high performance benefits from recent advances in random tensor-network contraction for the purposes of quantum circuit simulations~\cite{Gray2021, Pan2021, Kalachev2021}.
People developed algorithms to optimize the contraction order for tensor networks with thousands of vertices,
and the optimized contraction order gives us a near optimal space complexity $2^{\sim{\rm tw}(G)}$ to contract the tensor network.
For cases when the tensors being too large to fit into a GPU, slicing technique can be used to further reduce the space consumption.
Lastly, we provide a few examples and show that this toolbox can be used to calculate the hard-square entropy constant and overlap gap property. 
Our method can also be used for finding ``maximal'' independent set properties.
Here, the term ``maximal'' has a different meaning from ``maximum''  which we discussed above;
a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.
We show how to compute properties related to maximal independent sets as well as other combinatorial optimization problems in \App{app:otherproblems}.

\section{Tensor networks}
Tensor network~\cite{Cirac2020, Orus2014} is also known as einsum, factor graph or sum-product network~\cite{Bishop2006} in different contexts,
it can be viewed as a generalization of matrix multiplication to multiple tensors contraction.
People often use Einstein's notation to represent a tensor network, e.g. the matrix multiplication between two matrices $A$ and $B$ can be represented in Einstein's notation as $C_{ik} = A_{ij}B_{jk}$,
where we use a label in subscripts to represent a degree of freedom.
We enumerate these degree of freedoms and accumulate the product of tensor elements to the output tensor.
In the standard notation of Einstein's summation or tensor network in physics, each index appears precisely twice.
Hence people can represent a tensor network as a simple graph,
where a tensor is a vertex and two tensors sharing the same label are connected.
In the main text, we do not have this restriction. An index can appear arbitrary times.
The graphical representation of our generlized tensor network is hypergraph, in which an edge (label) can be shared by an arbitrary number of vertices (tensors).
These two notations are equivalent in representation power because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors - a high dimensional equivalence of identity matrix.
However, introducing $\delta$ tensors can potentially increase the contraction complexity a lot. We illustrate this point in \App{app:tensorbad}.
\blue{Is it better to introduce the contraction orders here or in example~\ref{eg:tensorcontraction}?}

@@

\section{Independence polynomial}
\subsection{Independence polynomial}\label{sec:indpoly}
The independence polynomial is an important graph polynomial that contains the counting information of independent sets. It is defined as
\begin{equation}\label{eq:idpdef}
I(G, x) = \sum_{k=0}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in $G$. The total number of independent sets is thus equal to $I(G, 1)$.
To compute the independence polynomial of a graph $G = (V, E)$, we reinterpret this problem as a tensor network contraction problem.
We map a vertex $i\in V$ to a label $s_i \in \{0, 1\}$ of dimension $2$ in a tensor network, where we use $0$ ($1$) to denote a vertex absent (exists) in the set.
For each label $s_i$, we defined a prametrized rank-one vertex tensor $W(x_i)$ indexed by it as
\begin{equation}
    W(x_i) = \left(\begin{matrix}
        1 \\
        x_i
    \end{matrix}\right).
\end{equation}
We use subscripts to index tensor elements, e.g. $W(x_i)_0=1$ is the first element associated with $s_i=0$ and $W(x_i)_1=x_i$ is the second element associated with to $s_i=1$.
Similarly, on each edge $(u, v)$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right). \label{eq:edgetensor}
\end{equation}
The corresponding tensor network contraction gives
\begin{equation}\label{eq:idp}
    P(G, \{x_1, x_{2}, \ldots,x_{|V|}\}) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{i=1}^{|V|} W(x_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j},
\end{equation}
where the summation runs over all vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$ (see Example \ref{eg:tensorcontraction} for a concrete example).
The edge tensor element $B_{s_{i}=1, s_{j}=1} = 0$ encodes the independent set constraint, meaning vertex $i$ and $j$ cannot be both in the independent set if they are connected by an edge $(i,j)$.
In the special case of $x_i = x$, the contraction result directly corresponds to the independence polynomial.
The connection can be understood as follows: the product over vertex tensor elements produces a factor $x^k$, where $k=\sum_i s_i$ counts the set size,
and the product over edge tensor elements gives a factor $1$ for a configuration being in an independent set and $0$ otherwise.

%In the implementation of the algorithm, we benefit from recent advances in tensor network contraction,
%where researchers working on quantum circuit simulation evaluate the tensor network by pairwise contracting tensors in a heuristic order~\cite{Gray2021, Pan2021}.
To evalute \Eq{eq:idp}, summing up the products directly is apparently computational inefficient.
The standard approach to evaluate a tensor network is to contract two tensors a time with a certain order utilizing the associativity and commutativity of tensor elements.
A good contraction order can reduce the time complexity significantly, at the cost of having a space overhead of $O(2^{\text{tw}(G)})$~\cite{Markov2008}.
The pairwise tensor contraction also makes it possible to utilize basic linear algebra subprograms (BLAS) functions to speed up the computation for certain tensor element types.

@@

@@

This set algebra can serve as the coefficients in \Eq{eq:countingtropical} to enumerate all MISs, \Eq{eq:polynomial} to enumerate independent sets of different sizes,
or \Eq{eq:max2poly} to enumerate all independent sets of size $\alpha(G)$ and $\alpha(G)-1$.
As long as the coefficients in a truncated polynomial forms a commutative semiring, the polynomial itself is a commutative semiring.
For example, to enumerate only the MISs, with the tropical algebra, we define $s_{k}\infty^k$,
where the coefficient follows the algebra in \Eq{eq:set} and the orders follows the max-plus tropical algebra.
The combined operations become: 
%\begin{equation}
%\begin{split}
%    s(k) \oplus s(j) &= \begin{cases}
%        \left(s_k \cup s_j, \max(k,j) \right), & k = j \\
%        \left(s_j, \max(k,j) \right), & k < j \\
%        \left(s_k, \max(k,j) \right), & k > j
%    \end{cases}, \\
%    s(k) \odot s(j) &= (\{\sigma \lor^\circ \tau \, | \, \sigma \in s_k, \tau \in s_j\}, k+j) \\
%    \mymathbb{0} &= (\{ \}, -\infty)\\
%    \mymathbb{1} &= (\{0^{\otimes |V|}\}, 0). \label{eq:countingtropicalset}
%\end{split}
%\end{equation}
\begin{equation}
\eqname{P1+SN}
\begin{split}
    s_x\infty^x \oplus s_y\infty^y &= \begin{cases}
        (s_x \cup s_y)\infty^{\max(x,y)}, & x = y\\
        s_y\infty^{\max(x,y)}, & x < y\\
        s_x\infty^{\max(x,y)}, & x > y
    \end{cases},\\
    s_x\infty^x \odot s_y\infty^y &= \{\sigma \lor^\circ \tau | \sigma \in s_x, \tau \in s_y\}\infty^{x+y},\\
    \mymathbb{0} &= \{\}\infty^{-\infty},\\
    \mymathbb{1} &= \{0^{\otimes |V|}\}\infty^{0}. \label{eq:countingtropicalset}
\end{split}
\end{equation}
Clearly, the vertex tensor and edge tensor become
\begin{equation}
    W^{\rm MISenum}(\{\boldsymbol{e}_{i}\}) = \left(\begin{matrix}
        \mymathbb{1} \\
        \{\boldsymbol{e}_{i}\}\infty^1
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm MISenum} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}
The contraction of the corresponding tensor network yields an enumeration of all MIS configurations.

@@

When we use the algebra in \Eq{eq:countingtropicalset} to enumerate all MIS configurations, we find that the program stores significantly more intermediate configurations than necessary and thus incur significant overheads in space.
To speed up the computation and reduce space overhead, we bound the searching space using the MIS size $\alpha(G)$.
As shown in \Fig{fig:bounding}, (a) we first compute the value of $\alpha(G)$ with tropical algebra and cache all intermediate tensors.
(b) Then, we compute a boolean mask for each cached tensor, where we use a boolean true to represent a tensor element having a contribution to the MIS (i.e.\ with a non-zero gradient) and boolean false otherwise.
(c) Finally, we perform masked tensor network contraction using the element type with the algebra in \Eq{eq:countingtropicalset} to obtaining all configurations.
Note that these masks in fact correspond to tensor elements with non-zero gradients with respect to the MIS size; we compute these masks by back propagating the gradients.
To derive the back-propagation rule for tropical tensor contraction,
we first reduce the problem to finding the back-propagation rule of a tropical matrix multiplication $C = A B$.
Since $ C_{ik} = \bigoplus_{j} \ A_{ij} \odot B_{jk} = \max_{j} \ A_{ij} \odot B_{jk}$ with tropical algebra, we have the following inequality
\begin{equation}
    A_{ij} \odot B_{jk} \leq C_{ik}.
\end{equation}
Here $\leq$ on tropical numbers are the same as the real-number algebra.
The equality holds for some $j'$, which means $A_{ij'}$ and $B_{j'k}$ have contributions to $C_{ik}$.
Intuitively, one can use this relation to indentify elements with nonzero gradients in $A$ and $B$ directly,
however, doing this might make us lost the advantage of using BLAS libraries~\cite{TropicalGEMM}.
Since $A_{ij} \odot B_{jk} = A_{ij} + B_{jk}$, one can move $B_{jk}$ to the right hand side of the inequality: 
\begin{equation}
    A_{ij} \leq C_{ik} \odot B_{jk}^{\circ -1}
\end{equation}
where ${}^{\circ -1}$ is the element-wise multiplicative inverse on tropical algebra (which is the additive inverse on real numbers).
The inequality still holds if we take the minimum over $k$: 
\begin{equation}
    A_{ij} \leq \min_{k}(C_{ik} \odot B_{jk}^{\circ -1}) = \left(\max_{k} \left(C_{ik}^{\circ -1} \odot B_{jk} \right) \right)^{\circ -1} = \left(\bigoplus_{k} \left(C_{ik}^{\circ -1} \odot B_{jk} \right) \right)^{\circ -1} = \left( C^{\circ-1} B^{\mathsf{T}} \right)^{\circ -1}_{ij}.
\end{equation}
On the right hand side, we transform the operation into a tropical matrix multiplication so that we can utilize the fast tropical BLAS routines~\cite{TropicalGEMM}.
Again, the equality holds if and only if the element $A_{ij}$ has a contribution to $C$ (i.e.\ having a non-zero gradient).
Let the gradient mask for $C$ be $\overline C$; the back-propagation rule for gradient masks reads
\begin{equation}\label{eq:adrule}
\overline{A}_{ij} = \delta \left(A_{ij}, \left( \left( C^{\circ-1} \circ \overline C \right) B^{\mathsf{T}} \right)_{ij}^{\circ -1} \right),
\end{equation}
where $\circ$ is the element-wise product, boolean false is treated as the tropical number $\mymathbb{0}$, and boolean true is treated as the tropical number $\mymathbb{1}$.
This rule defined on matrix multiplication can be easily generalized to tensor contraction by replacing the matrix multiplication between $C^{\circ-1} \circ \overline C$ and $B^{\mathsf{T}}$ by a tensor contraction.
With the above method, one can significantly reduce the space needed to store the intermediate configurations by setting the tensor elements masked false to zero during contraction.

>>>>

\begin{abstract}
We introduce a method of using generic programming tensor network to compute various properties of independent sets,
which includes: maximum size of such sets, the number of sets of a given size, enumeration of sets of a given size.
%Although the computational complexity inevitably scales exponentially with treewidth of the graph,
%our method using tensor network contraction with generic programming is highly versatile and is able to compute properties that are otherwise not known to be possible for up to several hundred vertices.
At the same time, our algorithms are very simple to implement and one can directly utilize recent advances in tensor network contraction techniques like near optimal contraction order finding and slicing.
%One can easily find a good contraction order for the tensor network representation of a graph $G$ with thousands of vertices. 
The algorithmic complexity of this approach is $2^{{\rm tw}(G)}$, where ${\rm tw} (G)$ the tree width of the problem graph.
Our generic programming tensor network method can be used to compute properties of other problems like cut size, colouring, maximal cliques et al.
To demonstrate the versatility of this tool, we apply it to a few examples including the calculations of the hard-square entropy constant and overlap gap property of three regular graphs.
\end{abstract}

@@

\section{Introduction}
In graph theory and combinatorial optimization, there are many interesting hard computational problems concerning properties of independent sets.
%An independent set of a graph is a subset of its vertices that no two vertices in this set are adjacent to each other.
For an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, $(u, v) \not\in E$. %there is no edge connecting $u$ and $v$ in $G$.
The primary interest of many computational complexity theorists is finding the maximum size of such vertex sets $\alpha(G) \equiv \max_{I}|I|$ and one of the sets of this size.
Finding $\alpha(G)$ is a hard computational problem; moreover, it is NP-hard to even approximate $\alpha(G)$ within $|V|^{1-\epsilon}$ ~\cite{Hastad1996},
meaning there is unlikely a polynomial time algorithm to find an independent set with size $\alpha(G)/|V|^{1-\epsilon}$ for an arbitrarily small positive $\epsilon$.
Naive exhaustive search for an MIS requires computing time $O\left(2^{|V|} \right)$.
People developed more efficient exact algorithms for finding MISs, including the branching algorithms~\cite{Tarjan1977,Robson1986} and dynamic programming.
The branching algorithms can reduce the base of the exponential time scaling to, e.g., $1.1996^{|V|} {|V|}^{O(1)}$~\cite{Xiao2017}, 
while dynamic programming approaches~\cite{Courcelle1990, Fomin2013} works better for graphs with a small tree width $\text{tw}(G)$, and
these methods can produce algorithms of complexity $O(2^{\text{tw}(G)} \text{tw}(G) |V|)$.
People are eager to find better algorithms for solving the MIS problem;
not only for it has a wide range of applications in scheduling, logistics, social network analysis, bioinformatics, wireless networks and telecommunication,
map labelling, computer vision, etc.~\cite{Butenko2003, Wu2015}, but also for it is a well know NP-complete problem that can be mapped from
 many other important combinatorial optimization problems in polynomial time such as 3-satisfiability problem, the maximum clique problem and the maximum vertex cover problem~\cite{Moore2011}.

%more concretely, an maximal independent set of a graph $G$ corresponds to a maximum clique in the complement graph of $G$,
%and the complement vertex set of an MIS corresponds to a minimum vertex cover of the graph $G$.
%In this paper, the term ``maximal'' has a different meaning from ``maximum''  which we discussed above; a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.
In this paper, we do not limit our discussion to finding one of the maximum independent sets and its size.
There are many other interesting problems pertaining to independent sets, such as the number of independent sets of a given size
and enumerating independent sets at a give size or what we call computing \textit{properties}.
Some of these problems are of great interest in physics applications such as the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018} \red{cite experiment when it's ready};
they can, for example, be used to understand phase transitions, to analyse the overlap gap property~\cite{Gamarnik2013, Gamarnik2019} and to identify harder graphs in an ensemble of graphs \red{cite experiment}.
Finding the number of independent sets of a given size is equivalent to computing the coefficients of independence polynomial,
which is a useful graph characteristics that can related to Euler characteristics of the independence complex~\cite{Bousquet2008,Levit2009}, partition functions~\cite{Lee1952,Yang1952} et al.
Computing independence polynomials belongs to the complexity class \#P-hard 
and only exponential algorithms~\cite{Ferrin2014} are known to compute it exactly.
There are also some interesting works on efficiently approximating the independence polynomial~\cite{Harvey2018}, but in this work, we mainly focus on computing the independence polynomial exactly.
We introduce a tensor-network based framework to compute the various properties pertaining to independent sets.
We map the independent set problem into generic tensor network contraction with specially designed tensor element algebra.
Its algorithmic complexity is similar to that of dynamic programming, i.e. scales exponentially as ${\rm tw}(G)$.
%But the tensor network approach is highly versatile so different problems can be generically solved with minimal implementation efforts without loss of efficiency.
We benchmark our algorithms by computing various independent set properties on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance on sparse graphs.
Its high performance benefits from recent advances in random tensor-network contraction for the purposes of quantum circuit simulations~\cite{Gray2021, Pan2021, Kalachev2021}.
People developed algorithms to optimize the contraction order for tensor networks with thousands of vertices,
and the optimized contraction order gives us a near optimal space complexity $2^{\sim{\rm tw}(G)}$ to contract the tensor network.
For cases when the tensors being too large to fit into a GPU, slicing technique can be used to further reduce the space consumption.
Lastly, we provide a few examples and show that this toolbox can be used to calculate the hard-square entropy constant and overlap gap property. 
Our method can also be used for finding ``maximal'' independent set properties.
Here, the term ``maximal'' has a different meaning from ``maximum''  which we discussed above;
a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.
We show how to compute properties related to maximal independent sets as well as other combinatorial optimization problems in \App{app:otherproblems}.

\section{Tensor networks}
Tensor network~\cite{Cirac2020, Orus2014} is also known as einsum, factor graph or sum-product network~\cite{Bishop2006} in different contexts,
it can be viewed as a generalization of matrix multiplication to multiple tensors contraction.
People often use Einstein's notation to represent a tensor network, e.g. the matrix multiplication between two matrices $A$ and $B$ can be represented in Einstein's notation as $C_{ik} = A_{ij}B_{jk}$,
where we use a label in subscripts to represent a degree of freedom.
We enumerate these degree of freedoms and accumulate the product of tensor elements to the output tensor.
In the standard notation of Einstein's summation or tensor network in physics, each index appears precisely twice.
Hence people can represent a tensor network as a simple graph,
where a tensor is a vertex and two tensors sharing the same label are connected.
In the main text, we do not have this restriction. An index can appear arbitrary times.
The graphical representation of our generalized tensor network is hypergraph, in which an edge (label) can be shared by an arbitrary number of vertices (tensors).
These two notations are equivalent in representation power because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors - a high dimensional equivalence of identity matrix.
However, introducing $\delta$ tensors can potentially increase the contraction complexity a lot. We illustrate this point in \App{app:tensorbad}.
\blue{Is it better to introduce the contraction orders here or in example~\ref{eg:tensorcontraction}?}

@@

\section{Independence polynomial}
\subsection{Independence polynomial}\label{sec:indpoly}
The independence polynomial is an important graph polynomial that contains the counting information of independent sets. It is defined as
\begin{equation}\label{eq:idpdef}
I(G, x) = \sum_{k=0}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in $G$. The total number of independent sets is thus equal to $I(G, 1)$.
To compute the independence polynomial of a graph $G = (V, E)$, we reinterpret this problem as a tensor network contraction problem.
We map a vertex $i\in V$ to a label $s_i \in \{0, 1\}$ of dimension $2$ in a tensor network, where we use $0$ ($1$) to denote a vertex absent (exists) in the set.
For each label $s_i$, we defined a parametrized rank-one vertex tensor $W(x_i)$ indexed by it as
\begin{equation}
    W(x_i) = \left(\begin{matrix}
        1 \\
        x_i
    \end{matrix}\right).
\end{equation}
We use subscripts to index tensor elements, e.g. $W(x_i)_0=1$ is the first element associated with $s_i=0$ and $W(x_i)_1=x_i$ is the second element associated with to $s_i=1$.
Similarly, on each edge $(u, v)$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right). \label{eq:edgetensor}
\end{equation}
The corresponding tensor network contraction gives
\begin{equation}\label{eq:idp}
    P(G, \{x_1, x_{2}, \ldots,x_{|V|}\}) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{i=1}^{|V|} W(x_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j},
\end{equation}
where the summation runs over all vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$ (see Example \ref{eg:tensorcontraction} for a concrete example).
The edge tensor element $B_{s_{i}=1, s_{j}=1} = 0$ encodes the independent set constraint, meaning vertex $i$ and $j$ cannot be both in the independent set if they are connected by an edge $(i,j)$.
In the special case of $x_i = x$, the contraction result directly corresponds to the independence polynomial.
The connection can be understood as follows: the product over vertex tensor elements produces a factor $x^k$, where $k=\sum_i s_i$ counts the set size,
and the product over edge tensor elements gives a factor $1$ for a configuration being in an independent set and $0$ otherwise.

%In the implementation of the algorithm, we benefit from recent advances in tensor network contraction,
%where researchers working on quantum circuit simulation evaluate the tensor network by pairwise contracting tensors in a heuristic order~\cite{Gray2021, Pan2021}.
To evaluate \Eq{eq:idp}, summing up the products directly is apparently computational inefficient.
The standard approach to evaluate a tensor network is to contract two tensors a time with a certain order utilizing the associativity and commutativity of tensor elements.
A good contraction order can reduce the time complexity significantly, at the cost of having a space overhead of $O(2^{\text{tw}(G)})$~\cite{Markov2008}.
The pairwise tensor contraction also makes it possible to utilize basic linear algebra subprograms (BLAS) functions to speed up the computation for certain tensor element types.

@@

% \begin{algorithm}[!ht]
%     \small
%     \SetAlgoNoLine
%     %\LinesNumbered
%     Let vector $\chi = (\chi_0,\chi_1,\chi_2, \ldots, \chi_{\alpha(G)})$ be a random sequence, matrix $X_{ij} = (\chi_i)^j$, where $i,j = 0, 1, \ldots, \alpha(G)$\;

%     \For{$i=0\ldots\alpha(G)$}{
%         $y_i$ = ${\rm contract\_tensor\_network}(\chi_i)$
%     }

%     \Return ${\rm gaussian\_elimination}(X, (y_0, y_1, \ldots, y_{\alpha{G}})) $\;
% \end{algorithm}

@@

This set algebra can serve as the coefficients in \Eq{eq:countingtropical} to enumerate all MISs, \Eq{eq:polynomial} to enumerate independent sets of different sizes,
or \Eq{eq:max2poly} to enumerate all independent sets of size $\alpha(G)$ and $\alpha(G)-1$.
As long as the coefficients in a truncated polynomial forms a commutative semiring, the polynomial itself is a commutative semiring.
For example, to enumerate only the MISs, with the tropical algebra, we define $s_{k}\infty^k$,
where the coefficient follows the algebra in \Eq{eq:set} and the orders follows the max-plus tropical algebra.
The combined operations become: 
%\begin{equation}
%\begin{split}
%    s(k) \oplus s(j) &= \begin{cases}
%        \left(s_k \cup s_j, \max(k,j) \right), & k = j \\
%        \left(s_j, \max(k,j) \right), & k < j \\
%        \left(s_k, \max(k,j) \right), & k > j
%    \end{cases}, \\
%    s(k) \odot s(j) &= (\{\sigma \lor^\circ \tau \, | \, \sigma \in s_k, \tau \in s_j\}, k+j) \\
%    \mymathbb{0} &= (\{ \}, -\infty)\\
%    \mymathbb{1} &= (\{0^{\otimes |V|}\}, 0). \label{eq:countingtropicalset}
%\end{split}
%\end{equation}
\begin{equation}
\eqname{P1+SN}
\begin{split}
    s_x\infty^x \oplus s_y\infty^y &= \begin{cases}
        (s_x \cup s_y)\infty^{\max(x,y)}, & x = y\\
        s_y\infty^{\max(x,y)}, & x < y\\
        s_x\infty^{\max(x,y)}, & x > y
    \end{cases},\\
    s_x\infty^x \odot s_y\infty^y &= \{\sigma \lor^\circ \tau | \sigma \in s_x, \tau \in s_y\}\infty^{x+y},\\
    \mymathbb{0} &= \{\}\infty^{-\infty},\\
    \mymathbb{1} &= \{0^{\otimes |V|}\}\infty^{0}. \label{eq:countingtropicalset}
\end{split}
\end{equation}
Clearly, the vertex tensor and edge tensor become
\begin{equation}
    W^{\rm MISenum}(\{\boldsymbol{e}_{i}\infty^1\}) = \left(\begin{matrix}
        \mymathbb{1} \\
        \{\boldsymbol{e}_{i}\}\infty^1
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm MISenum} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}
The contraction of the corresponding tensor network yields an enumeration of all MIS configurations.

@@

When we use the algebra in \Eq{eq:countingtropicalset} to enumerate all MIS configurations, we find that the program stores significantly more intermediate configurations than necessary and thus incur significant overheads in space.
To speed up the computation and reduce space overhead, we bound the searching space using the MIS size $\alpha(G)$.
As shown in \Fig{fig:bounding}, (a) we first compute the value of $\alpha(G)$ with tropical algebra and cache all intermediate tensors.
(b) Then, we compute a boolean mask for each cached tensor, where we use a boolean true to represent a tensor element having a contribution to the MIS (i.e.\ with a non-zero gradient) and boolean false otherwise.
(c) Finally, we perform masked tensor network contraction using the element type with the algebra in \Eq{eq:countingtropicalset} to obtaining all configurations.
Note that these masks in fact correspond to tensor elements with non-zero gradients with respect to the MIS size; we compute these masks by back propagating the gradients.
To derive the back-propagation rule for tropical tensor contraction,
we first reduce the problem to finding the back-propagation rule of a tropical matrix multiplication $C = A B$.
Since $ C_{ik} = \bigoplus_{j} \ A_{ij} \odot B_{jk} = \max_{j} \ A_{ij} \odot B_{jk}$ with tropical algebra, we have the following inequality
\begin{equation}
    A_{ij} \odot B_{jk} \leq C_{ik}.
\end{equation}
Here $\leq$ on tropical numbers are the same as the real-number algebra.
The equality holds for some $j'$, which means $A_{ij'}$ and $B_{j'k}$ have contributions to $C_{ik}$.
Intuitively, one can use this relation to identify elements with nonzero gradients in $A$ and $B$ directly,
however, doing this might make us lost the advantage of using BLAS libraries~\cite{TropicalGEMM}.
Since $A_{ij} \odot B_{jk} = A_{ij} + B_{jk}$, one can move $B_{jk}$ to the right hand side of the inequality: 
\begin{equation}
    A_{ij} \leq C_{ik} \odot B_{jk}^{\circ -1}
\end{equation}
where ${}^{\circ -1}$ is the element-wise multiplicative inverse on tropical algebra (which is the additive inverse on real numbers).
The inequality still holds if we take the minimum over $k$: 
\begin{equation}
    A_{ij} \leq \min_{k}(C_{ik} \odot B_{jk}^{\circ -1}) = \left(\max_{k} \left(C_{ik}^{\circ -1} \odot B_{jk} \right) \right)^{\circ -1} = \left(\bigoplus_{k} \left(C_{ik}^{\circ -1} \odot B_{jk} \right) \right)^{\circ -1} = \left( C^{\circ-1} B^{\mathsf{T}} \right)^{\circ -1}_{ij}.
\end{equation}
On the right hand side, we transform the operation into a tropical matrix multiplication so that we can utilize the fast tropical BLAS routines~\cite{TropicalGEMM}.
Again, the equality holds if and only if the element $A_{ij}$ has a contribution to $C$ (i.e.\ having a non-zero gradient).
Let the gradient mask for $C$ be $\overline C$; the back-propagation rule for gradient masks reads
\begin{equation}\label{eq:adrule}
\overline{A}_{ij} = \delta \left(A_{ij}, \left( \left( C^{\circ-1} \circ \overline C \right) B^{\mathsf{T}} \right)_{ij}^{\circ -1} \right),
\end{equation}
where $\circ$ is the element-wise product, boolean false is treated as the tropical number $\mymathbb{0}$, and boolean true is treated as the tropical number $\mymathbb{1}$.
This rule defined on matrix multiplication can be easily generalized to tensor contraction by replacing the matrix multiplication between $C^{\circ-1} \circ \overline C$ and $B^{\mathsf{T}}$ by a tensor contraction.
With the above method, one can significantly reduce the space needed to store the intermediate configurations by setting the tensor elements masked false to zero during contraction.
